---
title: "Customer Personality Analysis"
author: "Surya Eada"
output: pdf_document
---

# Introduction
This project aims to perform a brief analysis of customer data, shedding light on key insights and patterns within the dataset. Through visualization, we will uncover valuable information about customer demographics, purchase behavior, and relationships between variables. The analysis will involve visualizing data using R libraries like ggplot2, GGally, and Plotly, as well as observing correlation among various variables. This project seeks to provide actionable insights for marketing and business strategies, helping organizations make data-driven decisions and enhance their understanding of their target customers.

# Data Description
## Content
The dataset for this project, provided by Dr. Omar Romero-Hernandez, offers a comprehensive view of customer attributes, behaviors, and interactions with a company. It is designed to enable customer segmentation and enhance understanding of customer preferences and responses to marketing campaigns.

## Columns used from the dataset
Income : Income

Education : Level of Education

Marital_Status : Marital Status of the person

MntWines : Amount spent on wine in last 2 years

MntMeatProducts : Amount spent on meat products in last 2 years

MntGoldProds : Amount spent on gold products in last 2 years

MntFishProducts : Amount spent on Fish in last 2 years

MntSweetProducts : Amount spent on Sweet products in last 2 years

MntFruits : Amount spent on Fruits in last 2 years


# Importing Dataset

Here I'm importing the dataset that we will bw working on.
```{r Importing_Dataset}
df <- read.csv("Downloads/Customer_Personality_Analysis/marketing_campaign.csv",sep = "\t")
```


```{r Loading_Libraries,echo=FALSE,message=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(corrplot)
library(plotly)
library(tidyr)
library(lubridate)
library(GGally)
library(RColorBrewer)
library(gridExtra)
```

```{r Data_Cleaning,echo=FALSE}

# Check for missing values and calculate percentages
missing_values <- sapply(df, function(x) mean(is.na(x)) * 100)

# Drop rows with missing values
df <- na.omit(df)

# Print unique values of 'Education' column and 'Martial_Status' column
unique_education <- unique(df$Education)

unique_marital_status <- unique(df$Marital_Status)

# Create a copy of the original data frame
customer_data <- df

# Replace values in the 'Education' column and 'Martial_Status' column
replace_vals <- c('Graduation', 'Basic', '2n Cycle')
customer_data$Education <- ifelse(customer_data$Education %in% replace_vals, 'Undergraduate', 'Postgraduate')

replace_vals <- c('Together', 'Married')
customer_data$Marital_Status <- ifelse(customer_data$Marital_Status %in% replace_vals, 'In relationship', customer_data$Marital_Status)

replace_vals <- c("Alone", "Single")
customer_data$Marital_Status <- ifelse(customer_data$Marital_Status %in% replace_vals, 'Single', customer_data$Marital_Status)

# Create the "Age" column and define age bins and labels
current_year <- as.numeric(format(Sys.Date(), "%Y"))
customer_data$age <- current_year - customer_data$Year_Birth
age_bins <- c(0, 35, 45, 55, 65, 75, Inf)
age_labels <- c('Under 35', '35-45', '46-55', '56-65', '66-75', 'Over 75')

# Create the "age_group" column
customer_data$age_group <- cut(customer_data$age, breaks = age_bins, labels = age_labels, right = FALSE)

# Print unique values of 'Education' column and 'Martial Status' column
unique_education <- unique(customer_data$Education)

unique_marital_status <- unique(customer_data$Marital_Status)

# Remove unnecessary rows and columns from the data frame
columns_to_remove <- c('ID', 'Year_Birth', 'Z_CostContact', 'Z_Revenue', 'Complain')
customer_data <- customer_data[, !(names(customer_data) %in% columns_to_remove)]
customer_data <- customer_data[customer_data$Marital_Status != 'YOLO', ]
customer_data <- customer_data[customer_data$Marital_Status != 'Absurd', ]
customer_data <- customer_data[customer_data$Income <= 200000, ]
customer_data <- customer_data[customer_data$age <=100, ]
```


# Exploratory Data Analysis

Below, we shall sketch the densities of variables age and income as well as bargraphs to show the number of customers with each unique characteristics.

```{r EDA1,fig.height=3.9,fig.width=3.9,echo=FALSE}
# Create histograms separately for 'age' and 'Income'
plot_age <- ggplot(customer_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "coral", color = "black", aes(y = after_stat(density))) +
  geom_density(color = "coral4", linewidth = 1, alpha = 0.5) +  # Add density plot
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  theme_minimal()
print(plot_age)

plot_Income <- ggplot(customer_data, aes(x = Income)) +
  geom_histogram(binwidth = 5000, fill = "darkolivegreen2", color = "black", aes(y =after_stat(density))) +
  geom_density(color = "darkgreen", linewidth = 1, alpha = 0.5) +  # Add density plot
  labs(title = "Income Distribution", x = "Income", y = "Density") +
  theme_minimal()
plot_Income + scale_x_continuous(labels = scales::comma) + scale_y_continuous(labels = scales::comma)


# Distribution of categorical variables
ggplot(customer_data, aes(x = Education, fill = Education)) +
  geom_bar() +
  labs(title = "Education Distribution", x = "Education", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(customer_data, aes(x = Marital_Status, fill = Marital_Status)) +
  geom_bar() +
  labs(title = "Marital Status Distribution", x = "Marital Status", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The graphs below tell us about how the characterstics of the customer base of the company looks like to figure out the target customers for their marketing.

```{r EDA1.1,fig.height=3.9,fig.width=3.9,echo=FALSE}
# Relationships between variables
#scatter plot between 'Income' and 'age'

t1 <- ggplot(customer_data, aes(x = Income, y = age, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. Age", x = "Income", y = "Age") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set1")  # Use a color palette from RColorBrewer
t2 <- ggplot(customer_data, aes(x = Income, y = age, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. Age", x = "Income", y = "Age") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set2")  # Use a color palette from RColorBrewer
grid.arrange(t1,t2)
```

The below graphs plot income against the amount of money spent on various commodities offered by the company while also making clear of the characteristics of the customer base. These plots make it easier to see what the target customer base is for each commodity to better market the commodities.

```{r EDA2,fig.height=3.5,fig.width=3.5,echo=FALSE}
# Relationships between variables
#scatter plot between Income and Amounts spent on various places
p1 <- ggplot(customer_data, aes(x = Income, y = MntWines, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. MntWines", x = "Income", y = "MntWines") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set3")  # Use a color palette from RColorBrewer
p2 <- ggplot(customer_data, aes(x = Income, y = MntMeatProducts, color = Education)) +
  geom_point() +
  labs(title = "Income vs. MntMeatProducts", x = "Income", y = "MntMeatProducts") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set3")  # Use a color palette from RColorBrewer
p3 <- ggplot(customer_data, aes(x = Income, y = MntGoldProds, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. MntGoldProds", x = "Income", y = "MntGoldProds") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set3")  # Use a color palette from RColorBrewer
p4 <- ggplot(customer_data, aes(x = Income, y = MntFishProducts, color = Education)) +
  geom_point() +
  labs(title = "Income vs. MntFishProducts", x = "Income", y = "MntFishProducts") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set3")  # Use a color palette from RColorBrewer
p5 <- ggplot(customer_data, aes(x = Income, y = MntSweetProducts, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. MntSweetProducts", x = "Income", y = "MntSweetProducts") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set3")  # Use a color palette from RColorBrewer
p6 <- ggplot(customer_data, aes(x = Income, y = MntFruits, color = Education)) +
  geom_point() +
  labs(title = "Income vs. MntFruits", x = "Income", y = "MntFruits") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Set3")  # Use a color palette from RColorBrewer

q1 <- ggplot(customer_data, aes(x = Income, y = MntWines, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. MntWines", x = "Income", y = "MntWines") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Paired")  # Use a color palette from RColorBrewer
q2 <- ggplot(customer_data, aes(x = Income, y = MntMeatProducts, color = Marital_Status)) +
  geom_point() +
  labs(title = "Income vs. MntMeatProducts", x = "Income", y = "MntMeatProducts") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Paired")  # Use a color palette from RColorBrewer
q3 <- ggplot(customer_data, aes(x = Income, y = MntGoldProds, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. MntGoldProds", x = "Income", y = "MntGoldProds") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Paired")  # Use a color palette from RColorBrewer
q4 <- ggplot(customer_data, aes(x = Income, y = MntFishProducts, color = Marital_Status)) +
  geom_point() +
  labs(title = "Income vs. MntFishProducts", x = "Income", y = "MntFishProducts") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Paired")  # Use a color palette from RColorBrewer
q5 <- ggplot(customer_data, aes(x = Income, y = MntSweetProducts, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Income vs. MntSweetProducts", x = "Income", y = "MntSweetProducts") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Paired")  # Use a color palette from RColorBrewer
q6 <- ggplot(customer_data, aes(x = Income, y = MntFruits, color = Marital_Status)) +
  geom_point() +
  labs(title = "Income vs. MntFruits", x = "Income", y = "MntFruits") +
  scale_x_continuous(labels = scales::comma) +  # Format x-axis labels if needed
  scale_color_brewer(palette = "Paired")  # Use a color palette from RColorBrewer
grid.arrange(p1,q1)
grid.arrange(p2,q2)
grid.arrange(p3,q3)
grid.arrange(p4,q4)
grid.arrange(p5,q5)
grid.arrange(p6,q6)
```

The below graphs plot age against the amount of money spent on various commodities offered by the company while also making clear of the characteristics of the customer base. These plots make it easier to see what the target customer base is for each commodity to better market the commodities.

```{r EDA2.1,fig.height=3.5,fig.width=3.5,echo=FALSE}
r1 <- ggplot(customer_data, aes(x = age, y = MntWines, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntWines", x = "age", y = "MntWines") +
  scale_color_brewer(palette = "Accent")  # Use a color palette from RColorBrewer
r2 <- ggplot(customer_data, aes(x = age, y = MntMeatProducts, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntMeatProducts", x = "age", y = "MntMeatProducts") +
  scale_color_brewer(palette = "Accent")  # Use a color palette from RColorBrewer
r3 <- ggplot(customer_data, aes(x = age, y = MntGoldProds, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntGoldProds", x = "age", y = "MntGoldProds") +
  scale_color_brewer(palette = "Accent")  # Use a color palette from RColorBrewer
r4 <- ggplot(customer_data, aes(x = age, y = MntFishProducts, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntFishProducts", x = "age", y = "MntFishProducts") +
  scale_color_brewer(palette = "Accent")  # Use a color palette from RColorBrewer
r5 <- ggplot(customer_data, aes(x = age, y = MntSweetProducts, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntSweetProducts", x = "age", y = "MntSweetProducts") +
  scale_color_brewer(palette = "Accent")  # Use a color palette from RColorBrewer
r6 <- ggplot(customer_data, aes(x = age, y = MntFruits, color = Education)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntFruits", x = "age", y = "MntFruits") +
  scale_color_brewer(palette = "Accent")  # Use a color palette from RColorBrewer

s1 <- ggplot(customer_data, aes(x = age, y = MntWines, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntWines", x = "age", y = "MntWines") +
  scale_color_brewer(palette = "Dark2")  # Use a color palette from RColorBrewer
s2 <- ggplot(customer_data, aes(x = age, y = MntMeatProducts, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntMeatProducts", x = "age", y = "MntMeatProducts") +
  scale_color_brewer(palette = "Dark2")  # Use a color palette from RColorBrewer
s3 <- ggplot(customer_data, aes(x = age, y = MntGoldProds, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntGoldProds", x = "age", y = "MntGoldProds") +
  scale_color_brewer(palette = "Dark2")  # Use a color palette from RColorBrewer
s4 <- ggplot(customer_data, aes(x = age, y = MntFishProducts, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntFishProducts", x = "age", y = "MntFishProducts") +
  scale_color_brewer(palette = "Dark2")  # Use a color palette from RColorBrewer
s5 <- ggplot(customer_data, aes(x = age, y = MntSweetProducts, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntSweetProducts", x = "age", y = "MntSweetProducts") +
  scale_color_brewer(palette = "Dark2")  # Use a color palette from RColorBrewer
s6 <- ggplot(customer_data, aes(x = age, y = MntFruits, color = Marital_Status)) +
  geom_point() +
  labs(title = "Scatter Plot: Age vs. MntFruits", x = "age", y = "MntFruits") +
  scale_color_brewer(palette = "Dark2")  # Use a color palette from RColorBrewer
grid.arrange(r1,s1)
grid.arrange(r2,s2)
grid.arrange(r3,s3)
grid.arrange(r4,s4)
grid.arrange(r5,s5)
grid.arrange(r6,s6)

```

## THE CORRELATION MATRIX

The below table shows the correlation between various variables to better understand the customers.

```{r EDA3,fig.height=4,fig.width=7,echo=FALSE}
# Correlation matrix for numerical variables
numerical_vars <- select(customer_data, c("Income", "age", "MntWines","MntMeatProducts","MntGoldProds","MntFishProducts","MntSweetProducts","MntFruits"))
correlation_matrix <- cor(numerical_vars)
print(correlation_matrix) 
```

## THE PAIR PLOT and HEATMAP OF NUMERICAL VARIABLES
The below pair plot and the heatmap helps us understand the correlation between the variables much easier

```{r EDA3.1,fig.height=4,fig.width=7,echo=FALSE}


# Create a pairs plot of the numerical variables.
ggpairs(numerical_vars,
        aes(alpha = 0.01))


# Correlation heatmap
corrplot(correlation_matrix, method = "color")

```


# Results

1. The majority of customers are between the ages of 40 and 60.
2. The majority of customers earn between 25000 and 75000.
3. The majority of customers are in a relationship.
4. Less customers have finished their postgraduation in comparision to the customers who only completed their undergraduation.
5. We can observe that Post graduates spend more on wine and meat than on fish,fruits,sweets, and gold.
6. We can also observe that Post graduates over 60 years old tend to buy more than Post graduates below 60.
7. Customers with Income over 25000 seems to be spending considerably higher compared to ones below 25000.
8. Wine and Meat seems to be the major contributors in terms of amount spent.
9. Meat seems to have a significant correspondance with all other commodities except gold, while Wine seems to have a significant correspondance with only meat.

# Conclusion

It appears evident that the company should primarily target individuals in relationships who fall within the age range of 40 to 60. The company's marketing efforts should be concentrated on the income bracket of $25,000 to $75,000. However, if the company believes that the market for this income group is saturated, they could consider redirecting their marketing efforts towards either higher or lower income brackets. This could be achieved by offering a budget-friendly section to cater to customers with lower incomes or by promoting luxury commodities for those with higher incomes.

Additionally, the company should focus its marketing efforts on its popular products, such as wine and meat, unless the market for these products has reached its saturation point. In such a case, they should shift their attention to fish, sweets, and fruits, which exhibit a significant correlation with other commodities. However, there is no urgent need to emphasize gold yet, as the amount spent on gold by customers is nearly equal to that of other commodities, while its correlation with other commodities is relatively lower.
